Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-21 paper.tex
Updates on MAT etc.

Motivated by the large numbers of SARS-CoV-2 genomes currently
available, \cite{turakhia2021ultrafast} propose a similar approach
to encoding variation data for non-recombining organisms,
where a Newick tree is annotated by mutations~\citep{mcbroome2021daily}.
Similarly, the Phastsim~\citep{demaio2021phastsim} mutation simulator,
developed primarily to support simulation of SARS-CoV-2 trees
with hundreds of thousands of leaves, can optionally output a
Newick tree, where nodes are annotated with the mutations.
However, the dependence on the Newick format for encoding
topological information means that it suffers from all the
problems of newick [what?], and there is no software library support
[confirm].

>>>>

The advantages of storing mutations rather than sequences have
recently become clear in the context of vast genomic
datasets produced during the SARS-CoV-2 pandemic.
\cite{turakhia2021ultrafast} propose the
Mutation Annotated Tree (MAT)
format~\citep{mcbroome2021daily}---consisting of a Newick tree
and associated mutations in a binary format---as an efficient way to store
and process large viral datasets,
and achieve excellent compression and processing performance.
Similarly, \texttt{phastsim}~\citep{demaio2021phastsim}
was developed to simulate sequence evolution on such large SARS-CoV-2
phylogenies, and also outputs a tree annotated with mutations (not
in MAT format, however) to avoid the bottleneck of generating
and storing the simulated sequences.
While these methods illustate the advantages of the general
approach of storing ancestry and mutations over sequences,
they do not generalise beyond their immediate settings, and
no software library support is available.

% on the other hand, tskit is awesome.
Sites and mutations are fully integrated into the tree sequence data
model and the \tskit\ library.
As illustrated in Fig~\ref{fig-ts-example},
each site corresponds to some position
along the genome, and each mutation occurs at a given site, above
a particular node, with a given derived state (along with other
information, such as the time the mutation occured).
% Now talk about why tskit allows us to easily generate mutations
% on any input tree sequence topology.
